A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-05-17 inst/manuscript/manuscript.Rmd
minor revision updates

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches in a finite population spatial context.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between design-based inference and model-based inference is small when GRTS samples are used but large when SRS samples are used, suggesting that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\eqref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), isotropic (independent of direction), and decays with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$. @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics] -- one example is the exponential covariance function. @cressie1993statistics provides a thorough list of spatial covariance functions. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent (correlated) variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent (not correlated) variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is called the partial sill, $\sigma^2_{2}$ is called the nugget, and $\phi$ is called the range. We denote $\bm{\theta}$ as the vector of covariance parameters that composes $\bm{\delta}$. In Equation$~$\ref{equation:expcov}, $\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}$.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood (REML) [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\theta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\theta}}_{reml}$, the REML estimates of $\bm{\theta}$. Then $\bm{\hat{\beta}}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\theta}}_{reml}$.

@@

In this section we describe how we used simulated and real data to investigate performance between simple random sampling (SRS) and GRTS sampling as well as performance between design-based (DB) and model-based (MB) inference. In SRS and GRTS sampling, all population units had equal inclusion probabilities and were selected without replacement. The important distinction between SRS and GRTS is that SRS ignores spatial locations while sampling but GRTS explicitly incorporates them. Together, the two sampling plans (SRS and GRTS) combined with the two inference approaches (DB and MB) yielded four sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and the commonly-used SRS variance formula [@sarndal2003model; @lohr2009sampling] was used to estimate the variance. This variance formula is given by
\begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation}
where $z_i$ is the $i$th response value, $\bar{z}$ is the mean of all $z_i$, $n$ is the sample size, $N$ is the population size, and $f = (1 - n / N)$ ($f$ is often called the finite population correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used to estimate means and the local neighborhood variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate means and variances using restricted maximum likelihood. SRS, GRTS sampling, and design-based inference were implemented using the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey]. FPBK was implemented using the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal].

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. We then used real data from the 2012 National Lakes Assessment to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. With the real data, we were in control of only the random sampling process.

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{Histogram of a realized population for the normal response.}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{Histogram of a realized population for the skewed response.}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of realized populations simulated for the normal and skewed resposnes using the random layout and 50\% DRE.}
\label{fig:sim_pops}
\end{figure}

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. Within each simulation scenario and trial, SRS and GRTS samples were selected and then design-based and model-based inferences were used to estimate (design-based) or predict (model-based) the realized mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With model-based inference, covariance parameters and the realized mean were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias is taken as the average deviation between each trial's estimated (or predicted) mean ($\hat{\mu}_i$) and its realized mean ($\mu_i$): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes simulation trials. Root-mean-squared error is taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage is taken as the proportion of simulation trials where the realized mean is contained in its 95% confidence (or prediction) interval. These intervals are constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the conterminous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the halved average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

@@

We studied performance of the four sampling-inference combinations by selecting 2000 random SRS and GRTS samples of size $n = 50$, $n = 100$, and $n = 200$ from the realized ZMMI and Hg ppb populations and then analyzing the samples using MB and DB inference. In total, there were six separate scenarios (two responses crossed with three sample sizes). Within each SRS and GRTS sample, design-based and model-based inferences were used to estimate or predict the population mean and construct 95% coverage intervals. With model-based inference, the exponential covariance was assumed, and covariance parameters and the population mean were estimated using REML (separately for each SRS and GRTS sample). We used the same evaluation metrics as for the simulated data: mean bias, RMSE, and interval coverage. Mean bias is taken as the average deviation between each sample's estimated (or predicted) mean ($\hat{\mu}_i$) and the population mean ($\mu$) (of ZMMI or Hg ppb): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)$, where $i$ indexes simulation trials. Root-mean-squared error is taken as the square root of the average squared deviation between each sample's estimated (or predicted) mean and its population mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}$. Interval coverage is taken as the proportion of simulation trials where the population mean is contained in its 95% confidence (or prediction) interval. These intervals are constructed using the normal distribution.

@@

95% interval coverage of both ZMMI and Hg ppb for all four sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}. When $n = 50$, interval coverage for both responses is too low, though interval coverage is higher for ZMMI (symmetric response) than for Hg ppb (skewed response). When $n = 100$, ZMMI interval coverage is approximately 95% except for GRTS-DB, which has coverage around 92%, while Hg ppb interval coverage ranges from approximately 90% (GRTS-DB) to 93% (GRTS-MB). When $n = 200$, ZMMI interval coverage is approximately 95% while Hg ppb interval coverage ranges from approximately 93% (GRTS-DB) to 95% (GRTS-MB).  As with the simulated data, coverages for the NLA data tend to increase with the sample sizes, coverages tend to be higher for symmetric responses than for skewed responses, and the local neighborhood variance was slightly too small for small $n$, yielding slightly lower interval coverages than the other sampling-inference combinations. Recall that model-based inference defines interval coverage properties across realized populations. With the simulated data, we evaluated interval coverage across realized populations, but for the NLA data, we evaluated interval coverage within a single realized population for different samples. We did find that model-based coverages were similar to the design-based coverages, however, suggesting that for some realized populations it is reasonable to heuristically view data from separate random samples as being from approximately separate realized populations. But generally, if model-based intervals constructed from many random samples of a single realized population show improper coverage, this does not necessarily imply a deficiency in model-based inference. Tables for interval coverage in all six simulation scenarios are provided in the supporting information.

@@

The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. Design-based approaches rely on random sampling to estimate population parameters. Model-based approaches rely on distributional assumptions to predict realized values of a data-generating stochastic process. Though model-based approaches do not rely on random sampling, it can still be beneficial as a way to guard against preferential sampling. While design-based and model-based approaches have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them for finite population spatial data while implementing GRTS sampling and the local neighborhood variance estimator. Aside from the theoretical differences described throughout the manuscript, a few analytical findings from the simulated and real data studies were particularly notable. All sampling-inference combinations had approximately zero mean bias. Independent of the inference approach, GRTS-DB and GRTS-MB had lower RMSE than their SRS counterparts. Though GRTS-DB and GRTS-MB generally had very similar RMSE, SRS-MB tended to have much lower RMSE than SRS-DB, suggesting that the model-based inference mitigated much of the inefficiency in RMSE from SRS. As the proportion of dependent random error in the simulated data increased, SRS-MB, GRTS-DB, and GRTS-MB become increasingly more efficient (lower RMSE) than SRS-DB. Interval coverage tended to be higher for the symmetric responses than skewed responses and tended to increase with the sample size. At a sample size of $n = 200$, generally all interval coverages were near the desired value of 95%.

@@

Now, we discuss advantages of the model-based approach. The model-based approach can more naturally quantify the relationship between covariates (predictor variables) and the response variable than design-based approaches. Model-based inference also yields estimated spatial covariance parameters, which help better understand the dependence structure of the process in study. Model selection is also possible using model-based inference and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based inference is capable of more efficient small-area estimation than design-based inference because model-based inference can leverage distributional assumptions in areas with few observed population units. Model-based approaches also accommodate unit-by-unit predictions at unobserved locations that can be used to construct informative visualizations like smoothed maps. @brus1997random provide a more thorough discussion regarding the benefits and drawbacks of the two approaches. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

@@

We would like to thank the editors and anonymous reviewers for hard work and time spent providing us with thoughtful, valuable feedback which greatly improved the manuscript.

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches in a finite population spatial context.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest, and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between design-based inference and model-based inference is small when GRTS samples are used but large when SRS samples are used, suggesting that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\eqref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), and isotropic (independent of direction) [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$. @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics] -- one example is the exponential covariance function. @cressie1993statistics provides a thorough list of spatial covariance functions. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent (correlated) variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent (not correlated) variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is called the partial sill, $\sigma^2_{2}$ is called the nugget, and $\phi$ is called the range. We denote $\bm{\theta}$ as the vector of covariance parameters that composes $\bm{\delta}$. In Equation$~$\ref{equation:expcov}, $\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}$.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on restricted maximum likelihood (REML) [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\theta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\theta}}_{reml}$, the REML estimates of $\bm{\theta}$. Then $\bm{\hat{\beta}}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\theta}}_{reml}$.

@@

In this section we describe how we used simulated and real data to investigate performance between simple random sampling (SRS) and GRTS sampling as well as performance between design-based (DB) and model-based (MB) inference. In SRS and GRTS sampling, all population units had equal inclusion probabilities and were selected without replacement. The important distinction between SRS and GRTS is that SRS ignores spatial locations while sampling but GRTS explicitly incorporates them. Together, the two sampling plans (SRS and GRTS) combined with the two inference approaches (DB and MB) yielded four sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and the commonly-used SRS variance formula [@sarndal2003model; @lohr2009sampling] was used to estimate variances. This variance formula is given by
\begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation}
where $z_i$ is the response value of the $i$th sampled population unit, $\bar{z}$ is the mean of all $z_i$, $n$ is the sample size, $N$ is the population size, and $f = (1 - n / N)$ ($f$ is often called the finite population correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used to estimate means and the local neighborhood variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate means and variances using restricted maximum likelihood. SRS, GRTS sampling, and design-based inference were implemented using the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey]. FPBK was implemented using the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal].

The simulated and real data were used in distinct ways to answer distinct questions. We used the simulated data to compare the sampling-inference combinations across many realized populations (from the same data-generating stochastic process) and random samples. We used the real data to compare the sampling-inference combinations within a single realized population but across random samples. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. With the real data, we were only in control of the random sampling process, which is typically the case in reality.

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{Histogram of a single realized population for the normal response.}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{Histogram of a single realized population for the skewed response.}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of single realized populations simulated for the normal and skewed respones using the random layout and 50\% DRE.}
\label{fig:sim_pops}
\end{figure}

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. First, a population was simulated according to the specifications of the particular simulation scenario (for the random location layout, locations were simulated separately for each trial). Next, SRS and GRTS samples were selected. Then, design-based and model-based inferences were used to estimate (design-based) or predict (model-based) the realized mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With model-based inference, covariance parameters and the realized mean were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias was taken as the average deviation between each trial's estimated (or predicted) mean ($\hat{\mu}_i$) and its realized mean ($\mu_i$): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes the simulation trials. Because each trial had different realized populations, $\mu_i$ changed with $i$. Root-mean-squared error was taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage was taken as the proportion of simulation trials where the realized mean was contained in its 95% confidence (or prediction) interval. These intervals were constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the conterminous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the halved average squared differences (semivariance) of responses whose separation (distance) falls within a separation class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

@@

We studied performance of the four sampling-inference combinations by selecting 2000 SRS and GRTS samples of size $n = 50$, $n = 100$, and $n = 200$ from the realized ZMMI and Hg ppb populations and then analyzing the samples using MB and DB inference. In total, there were six separate scenarios (two responses crossed with three sample sizes). Within each SRS and GRTS sample, design-based and model-based inferences were used to estimate or predict the population mean and construct 95% coverage intervals. With model-based inference, the exponential covariance was assumed, and covariance parameters and the population mean were estimated using REML (separately for each SRS and GRTS sample). We used the same evaluation metrics as for the simulated data: mean bias, RMSE, and interval coverage. Mean bias was taken as the average deviation between each sample's estimated (or predicted) mean ($\hat{\mu}_i$) and the population mean ($\mu$) (of ZMMI or Hg ppb): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)$, where $i$ indexes the simulation trials. Because each trial had the same realized population, $\mu$ did not change with $i$ (in contrast to the simulated data, where the realized mean changed with $i$). Root-mean-squared error was taken as the square root of the average squared deviation between each sample's estimated (or predicted) mean and its population mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}$. Interval coverage was taken as the proportion of simulation trials where the population mean was contained in its 95% confidence (or prediction) interval. These intervals were constructed using the normal distribution.

@@

95% interval coverage of both ZMMI and Hg ppb for all four sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}. When $n = 50$, interval coverage for both responses is too low, though interval coverage is higher for ZMMI (symmetric response) than for Hg ppb (skewed response). When $n = 100$, ZMMI interval coverage is approximately 95% except for GRTS-DB, which has coverage around 92%, while Hg ppb interval coverage ranges from approximately 90% (GRTS-DB) to 93% (GRTS-MB). When $n = 200$, ZMMI interval coverage is approximately 95% while Hg ppb interval coverage ranges from approximately 93% (GRTS-DB) to 95% (GRTS-MB).  As with the simulated data, coverages for the NLA data tend to increase with the sample sizes, coverages tend to be higher for symmetric responses than for skewed responses, and the local neighborhood variance was slightly too small for small $n$, yielding slightly lower interval coverages than the other sampling-inference combinations. Recall that model-based inference defines interval coverage properties across realized populations. With the simulated data, we evaluated interval coverage across realized populations, but for the NLA data, we evaluated interval coverage within a single realized population. We did find that model-based coverages were similar to the design-based coverages, however, suggesting that for some realized populations it is reasonable to heuristically view data from separate random samples as being from approximately separate realized populations. But generally, if model-based intervals constructed from many random samples of a single realized population show improper coverage, this does not necessarily imply a deficiency in model-based inference. Tables for interval coverage in all six simulation scenarios are provided in the supporting information.

@@

The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. Design-based approaches rely on random sampling to estimate population parameters. Model-based approaches rely on distributional assumptions to predict realized values of a data-generating stochastic process. Though model-based approaches do not rely on random sampling, random sampling can still be beneficial as a way to guard against preferential sampling. While design-based and model-based approaches have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them for finite population spatial data while implementing GRTS sampling and the local neighborhood variance estimator. Aside from the theoretical differences described throughout the manuscript, a few analytical findings from the simulated and real data studies were particularly notable. All sampling-inference combinations had approximately zero mean bias. Independent of the inference approach, the GRTS samples yielded lower RMSE than their SRS counterparts. Though GRTS-DB and GRTS-MB generally had very similar RMSE, SRS-MB tended to have much lower RMSE than SRS-DB, suggesting that the model-based inference mitigated much of the inefficiency in RMSE from SRS. As the proportion of dependent random error in the simulated data increased, SRS-MB, GRTS-DB, and GRTS-MB become increasingly more efficient (lower RMSE) than SRS-DB. Interval coverage tended to be higher for the symmetric responses than skewed responses and tended to increase with the sample size. At a sample size of $n = 200$, generally all interval coverages were near the desired value of 95%.

@@

Now, we discuss advantages of the model-based approach. The model-based approach can more naturally quantify the relationship between covariates (predictor variables) and the response variable than design-based approaches. Model-based inference also yields estimated spatial covariance parameters, which help better understand the dependence structure of the process in study. Model selection is also possible using model-based inference and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based inference is capable of more efficient small-area estimation than design-based inference because model-based inference can leverage distributional assumptions in areas with few observed population units. Model-based approaches also accommodate unit-by-unit predictions at unobserved locations that can be used to construct informative visualizations like smoothed maps. @brus1997random provide a more thorough discussion regarding the benefits and drawbacks of the two approaches. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, these benefits and drawbacks should be considered alongside the particular goals of the study.

@@

We would like to thank the editors and anonymous reviewers for their hard work and time spent providing us with thoughtful, valuable feedback which greatly improved the manuscript.
